---
title: "Age Curves"
subtitle: "Baseball Analytics Club: J-Term"
output: html_document
date: 'January, 2023'
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Open+Sans:wght@300&family=Source+Sans+Pro&display=swap" rel="stylesheet">
<style>

body{
font-family: 'Source Sans Pro', sans-serif;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
# load packages
library(Lahman)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
```

## _Choose a player:_
```{r, message=FALSE, warning=FALSE}
# firstName <- "Brandon"
# lastName<- "Belt"
# firstName <- "Xander"
# lastName<- "Bogaerts"
firstName <- "Carlos"
lastName<- "Correa"
```

```{r, message=FALSE, warning=FALSE}
# pull player id
batter_id <- People %>% 
  filter(nameFirst == firstName, nameLast == lastName) %>% 
  pull(playerID)

# Get chosen player's batting stats and calculate age for each season
batter <- battingStats() %>% 
  filter(playerID == batter_id) %>% 
  inner_join(People, by = "playerID") %>% 
  mutate(age = yearID - birthYear)

# Create player summary for comps
batter_summary <- batter %>% 
  summarize(debut = min(age), current = max(age), 
            TotalG = sum(G), 
            CareerAvg = round(sum(H) / sum(AB), 3),
            CareerOBP = round(sum(H + BB + HBP) / sum(AB + BB + HBP + SF), 3),
            CareerSLG = round(sum(H + X2B + 2*X3B + 3*HR) / sum(AB), 3),
            avgOPS = mean(OPS), highOPS = max(OPS)) %>%
  mutate(CareerISO = CareerSLG - CareerAvg,
         CareerOPS = CareerOBP - CareerSLG)
```

```{r, message=FALSE, warning=FALSE, fig.align='center'}
# Graph player's career to date
batter %>% 
  ggplot(aes(x = age, y = OPS)) +
  geom_point(aes(size = AB)) +
  geom_smooth() +
  geom_vline(xintercept = batter_summary$current, color = "red", size = 1) +
  geom_vline(xintercept = batter$age[batter$OPS == batter_summary$highOPS],
             color = "purple", size = 1, alpha = 0.5) +
  # added in a line to show which age had highOPS
  labs(title = paste(batter$nameFirst, batter$nameLast)) + 
  #scale_x_continuous(breaks = c(23:33)) + # updated axis to be whole numbers
  scale_x_continuous(breaks = c(min(batter$age):max(batter$age))) +
        # more universal way to fix the axis increments
  theme_bw()
```

<span style="color: darkblue;">
Updates from the 1/17 meeting:
</span>

### Comparison

#### _Pick the positions_

```{r}
positions <- c("2B", "SS", "3B")
```

```{r, message=FALSE, warning=FALSE, fig.align='center'}
# find players who played comparable position(s) as chosen player
fielders <- Fielding %>%
  filter(POS %in% positions, yearID > 1990)%>%
  group_by(playerID, yearID)%>%
  summarize(games = sum(G)) %>% 
  filter(games > 100) %>% 
  group_by(playerID) %>% 
  summarize(seasons = n()) %>% 
  filter(seasons > 2) %>% 
  ungroup

# add age
batting <- battingStats() %>% 
  inner_join(People, by = "playerID") %>% 
  mutate(age = yearID - birthYear)

# Find comps for chosen player - based on age, OPS - ranked by similarity
Comps <- batting %>% 
  group_by(playerID, nameFirst, nameLast) %>% 
  mutate(debut = min(age), reachedage = max(age) > batter_summary$current) %>%
  filter(debut < batter_summary$debut + 2, yearID > 1990, reachedage, 
         age < batter_summary$current + 2) %>%
  summarize(CareerOBP = round(sum(H + BB + HBP) / sum(AB + BB + HBP + SF), 3),
            CareerAvg = round(sum(H) / sum(AB), 3),
            CareerSLG = round(sum(H + X2B + 2*X3B + 3*HR) / sum(AB), 3),
            TotalAB = sum(AB), highOPS = max(OPS)) %>%
  mutate(CareerISO = CareerSLG - CareerAvg,
         CareerOPS = CareerOBP + CareerSLG,
         OBPDiff = round(CareerOBP - batter_summary$CareerOBP, 3),
         ISODiff = round(CareerISO - batter_summary$CareerISO, 3)) %>%
  filter(TotalAB > 1000, CareerOPS > batter_summary$CareerOPS - 0.050, 
         CareerOPS < batter_summary$CareerOPS + 0.050, 
         highOPS > batter_summary$highOPS - 0.050,
         highOPS < batter_summary$highOPS + 0.050) %>%
  ungroup() %>%
  inner_join(fielders, by = "playerID") %>%
  mutate(CompScore = abs(OBPDiff) + abs(ISODiff)) %>%
  arrange(CompScore)

# top comps
head(Comps)
nrow(Comps)

# find playerIDs for top 4 comps
example_id <- People %>% 
  slice_head(n = 4)%>% 
  pull(playerID)

# get their batting stats, calculate age, and add name
example <- battingStats() %>% 
  filter(playerID %in% example_id) %>% 
  inner_join(People, by = "playerID") %>% 
  mutate(age = yearID - birthYear, name = paste(nameFirst, nameLast))

# Graph comps for the chosen players
example %>% 
  ggplot(aes(x = age, y = OPS)) +
  geom_point(aes(size = G)) +
  geom_smooth(span = 1) +
  geom_vline(xintercept = batter_summary$current, color = "red", size = 1) +
  geom_vline(xintercept = batter$age[batter$OPS == batter_summary$highOPS],
             color = "purple", size = 1) +
  labs(title = paste("Comps for", firstName, lastName)) +
  #scale_x_continuous(breaks = c(min(example$age):max(example$age))) +
  theme_bw() +
  facet_wrap(vars(name))
# are we keeping the red line at 33 because that is how old Brandon Belt is,
# and we want to predict/get and idea of what his OPS will be like in coming
# years?
```

<span style="color: darkblue;">
From 1/3 meeting:
</span>

## Random player:
```{r, message=FALSE, warning=FALSE, fig.align='center'}
ids <- Batting %>% 
  group_by(playerID) %>% 
  mutate(seasons = n()) %>% 
  filter(seasons > 5, yearID > 1960) %>% 
  pull(playerID) %>% 
  unique()
# pull player id
example_id <- People %>% 
  filter(playerID == sample(ids, 1)) %>% 
  pull(playerID)

# get their batting stats
example <- battingStats() %>% filter(playerID == example_id)

# add in the age variable
example_update <- example %>% 
  inner_join(People, by = "playerID") %>% 
  mutate(age = yearID - birthYear)

example_update %>% 
  ggplot(aes(x = age, y = OPS)) +
  geom_point(aes(size = G)) +
  geom_smooth() +
  geom_vline(xintercept = 33, color = "red", size = 1) +
  labs(title = paste(example_update$nameFirst, example_update$nameLast),
       subtitle = paste("Career: ", min(example_update$yearID), " - ",
                        max(example_update$yearID))) +
  theme_bw()
```

### Joining data frames

Filtering battingStats to be group of interest.
```{r, message=FALSE, warning=FALSE}
base <- battingStats() %>%
  filter(yearID > 1990) %>%
  inner_join(People, by = "playerID") %>% 
  mutate(age = yearID - birthYear) %>%
  group_by(playerID, nameFirst, nameLast) %>% 
  mutate(debut = min(age), current = max(age), TotalG = sum(G), 
         avgOPS = mean(OPS), highOPS = max(OPS)) 
base <- base %>% 
  dplyr::select(-c(deathYear:deathCity, deathDate))
```

Merge with Fielding:
```{r, message=FALSE, warning=FALSE}
# filter variables
subFielding <- Fielding %>% 
  filter(yearID > 1990) %>% 
  dplyr::select(playerID, yearID, POS, GS, InnOuts, PO, A, E, DP)

basewf <- base %>% 
  inner_join(subFielding, by = c("playerID", "yearID"))
```

Add Salary:
```{r, message=FALSE, warning=FALSE}
subSal <- Salaries %>% 
  dplyr::select(-c(teamID, lgID)) #subset to take out repeated variables
basewf_ws <- basewf %>% 
  inner_join(subSal, by = c("playerID", "yearID"))
```

```{r, message=FALSE, warning=FALSE}
# reorder the variables
reordered <- basewf_ws %>% 
  subset(select = c(playerID:BABIP, nameGiven:debut, birthDate:highOPS,
                    salary, POS, birthYear:nameLast, GS:DP, finalGame:bbrefID))
```

#### Salary overall:
```{r, message=FALSE, warning=FALSE, fig.align='center'}
reordered %>% 
  ggplot(aes(x = age, y = salary)) +
  geom_smooth(se = FALSE, color = "red") +
  scale_y_continuous(labels = scales::comma) +
  theme_bw()
reordered %>% 
  ggplot(aes(x = salary)) +
  geom_smooth(aes(y = OPS)) +
  geom_smooth(aes(y = BABIP), color = "red") +
  annotate("text", label = "OPS", x = 31000000, y = 0.68, size = 5, 
           color = "blue") +
  annotate("text", label = "BABIP", x = 31000000, y = 0.3, size = 5, 
           color = "red") +
  scale_x_continuous(labels = scales::comma) +
  theme_bw() +
  theme(axis.title.y = element_blank())
```

#### Look at example player:
**The graphs were not working with Carlos Correa since the filtering greatly limited the number of observations. So the following code uses Brandon Belt as the example player.**

```{r, message=FALSE, warning=FALSE, fig.align='center'}
batter_all <- reordered %>% filter(playerID == "beltbr01")

batter_all %>% 
  ggplot(aes(x = age, y = salary)) +
  geom_point() +
  #geom_smooth() +
  geom_line() +
  labs(title = paste(batter_all$nameFirst, batter_all$nameLast)) + 
  scale_x_continuous(breaks = c(min(batter_all$age):max(batter_all$age))) + 
  # updated axis to be whole numbers
  theme_bw() +
  scale_y_continuous(labels = scales::comma) 

batter_all %>% 
  ggplot(aes(x = age)) +
  geom_smooth(aes(y = OPS)) +
  geom_smooth(aes(y = BABIP), color = "red") +
  labs(title = paste(batter_all$nameFirst, batter_all$nameLast)) + 
  #scale_x_continuous(breaks = c(min(batter_all$age):max(batter_all$age))) +
  annotate("text", label = "OPS", x = 23.8, y = 0.82, size = 5, color = "blue") +
  annotate("text", label = "BABIP", x = 23.8, y = 0.37, size = 5, 
           color = "red") +
  theme_bw()

# batter_all %>% 
#   ggplot(aes(x = OPS, y = salary)) +
#   geom_point(size = 2) +
#   geom_smooth() +
#   theme_bw() +
#   scale_y_continuous(labels = scales::comma) 
```

#### Graphs with position
```{r, message=FALSE, warning=FALSE, fig.align='center'}
reordered %>% 
  ggplot(aes(x = age, y = salary, color = POS)) +
  geom_smooth(se = FALSE) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma)
reordered %>% 
  ggplot(aes(x = age, y = OPS, color = POS)) +
  geom_smooth(se = FALSE) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma)
reordered %>% 
  ggplot(aes(x = OPS, y = salary, color = POS)) +
  geom_smooth(se = FALSE) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma)
```


